---
title: "Manipulating Data"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

### Load packages
Use the readr, tidyr and dplyr from Tidyverse. Load in all the packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(here)
```

### Practice with pivoting
Pivoting is described as a summary of the data, based on a certain category. You can pivot large datasets to provide a summary. The practical starts with a non tidy dataset

An important note is that for a dataset to be tidy each row must contain only one observation, in this dataset there is 3 observations per row. a1,b1 and c1 are all contained within one row
```{r}
patientID <- c(1,2)
test_result_month1 <- c("a1","a2")
test_result_month2 <- c("b1","b2")
test_result_month3 <- c("c1","c2")

patient_test <- data.frame(patientID, test_result_month1,test_result_month2,test_result_month3)
patient_test 
```
- The first thing we have to do is reshape the dataset, %>% is a pipe function that will lead one function into the other using the result of the previous function. we use the result of the first argument to pipe into the other function. We use the pivot longer function, this function will take a dataset that is wide in size and change it into something long. In this case we go from 3 observations in a row to 1, to do this we need to change the shape of the dataset. It does this by the user defining, new character and value columns
```{r}
tidy_patient_tests<- patient_test %>% 
    pivot_longer(
        c('test_result_month1','test_result_month2','test_result_month3'),
        names_to='month',
        values_to='test_result'
    )
tidy_patient_tests
```

To make the overall document look cleaner we can remove the long variable name and add a small prefix to it, to replace the variable name. The month value replaces all text. It takes all the text that is within 
```{r}
tidy_patient_tests <- patient_test %>%
  pivot_longer(
    c('test_result_month1','test_result_month2','test_result_month3'),
    names_to= 'month',
    names_prefix = 'test_result_month',
    names_transform = list(month = as.integer),
    values_to='test_result'
    )
tidy_patient_tests
```

### Reading in FEV data

Just read in the values for the FEV data:
```{r}
fev_data <- read_csv("C:\\Users\\romin\\ToyRepo\\fev.csv")
```

Using the $ symbol we can extract certain qualities about the data. The $ can have many functions within the code, but the one used here is to select a paticular item in the code. 
This result will be saved as a vector form, containing information about the data.  
```{r}
fev_data$fev[32] # Returns the 32nd element from the fev column
fev_data[32,3] # Returns the 32nd element of the 3rd column in fev 
fev_data[32, "age"] # Same as above, it just makes it more readable. 3 = "age" column
fev_data[32, ] #Everything in the 32nd element of fev column is returned
fev_data[32, 1:3] # Returns 32nd element from the 1st to the 3rd column
fev_data[32, -5] # This means that the index value 5 will be exculded from the vector
fev_data[32, -1:-2] # This excludes a slice of the dataset in the result 
```

To drop the 1st, 3rd and 5th column we do the following:
```{r}
fev_data[c(-1,-3,-5)]
```

### Logicals

Logicals such as TRUE or FALSE values can also be held in a vector. You can also set conditions to them such as restricting height
```{r}
is_tall <- fev_data$height > 72 # Takes the height column and restricts the data to values where height is greater than 72
```

To print these values we can use a summary table
```{r}
table(is_tall)
```

To be more specific/detailed of what the values actually result in we can pass the condition inside of the dataframe vector
```{r}
fev_data[is_tall,]
```