--- 
title: "Analysis of Temporal Trends in Cancer Registry Data in NHS Scotland "
output: 
  pdf_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


# Introduction
## 1.1 Background
Understanding trends within a countries cancer registry data provided an essential understanding for health budget allocation and informed public health polices,
targeted at reducing the incidence of certain cancers. Cancer diagnoses places significant strain on not only a patients life, but on the health care budget allocated. 
This form of stress on an economies healthcare budget is known as the cancer burden. The larger the cancer burden the harder the economy must work to meet the growing 
demands placed on the healthcare system to provide the patients with the intensive care required. To illustrate this problem we can examine how  
, Europe currently has the highest health expenditure
on cancer treatment also has the highest rate of deaths in a working age (30-69) due to cancer. Despite increased funding, quality healthcare and developing research into improving 
cancer care, providing the correct care in a timely manner that ensures cancer is caught earlier to prevent increased costs on a health care system still proves an issue. This is 
where maintaining and updating health statistics is important to inform policy makers on the changing needs amid rising cases. 


# Methods
## Data Loading/Cleansing
First the relevant library packages are installed, there are som typical boilerplate packages. Of note to this paticular report, are the packages:

- Lemon - provides functionalities for geom_pointline
- kableExtra - provides functionlities to make a table in R Markdown
- Zoo - provides mathematical functions
- ggrepel - provides functionalities to ensure that text labels dont overalp on graphs
### Load in Specific Packages
```{r Load Libraries}
suppressWarnings({
library(readr)
library(tidyr)
library(dplyr)
library(here)
library(lemon)
library(kableExtra)
library(ggplot2)
library(data.table)

library(zoo)
library(ggrepel)
})
```

Once all the relevant libraries are loaded in, the data is read in from a local directory and saved into a dataframe. In this case 
the dataframe that will be used for all data manipulation of the entire cancer registry data will be named *cancerReg* 

### Load in the data 
```{r Read in the Data}
cancerReg <- read.csv("C:\\Users\\romin\\ToyRepo\\Models\\cancerReg.csv")
cancerRegTable1 <- cancerReg[c(1,2,3,4,5,8)]
cancerRegTable2 <- cancerReg[c(6,9,10,11)]
kable(head(cancerRegTable1))
kable(head(cancerRegTable2))
```


Not all the data provided in the inital dataframe is required for further analysis and this line of code depicts the values in which we will not need.
These values have been removed to make the data cleaner and easier to manipulate. A notable exclusion is the "numerator" column, this column is the 
raw value of the amount of people under a specific health board in a specific time frame are contained in the cancer registry. This was due to the fact 
that the "numerator" values were not standardised across different health boards, and if the raw values were taken, it could lead to false conclusions over the data.

Especially with the variance in  population age between the healthboards, the largest difference being between Dumfries and Galloway with over 65's representing 29% of their population 
compared to Grampian's 20%. As noted in the Background section most new incidence of cancer come from an aging populaion. So for the sake of genralisability of results, a standardised value 
such as the measure is used throughout the interperetations and calulcations. The measure value itself is not standardised to a fine degree, but this will be further discussed in the results section

### Remove Uncessary Data for Analysis
```{r Remove Columns}
cancerReg <- cancerReg %>% select(
    -period, -area_type, -type_definition, -indicator,
    -upper_confidence_interval, -lower_confidence_interval, -numerator
)
```

## Visual Representation of All Values in Cancer Registry Data
To understand how to best analyse the data, a general overview is helpful to aid in this. This portion of the assignemnt took the longest 
since it was difficult to map for all 14 healthboards in a clear and conscise manner without cluttering the plot. especially when there such drastic changes in the measure 
variable between time periods. Initally, a variation of plots such as stacked area chart and density plots were experimented with but they all proved too limited in their 
visual represnations as overlapping data, and the jumping of values were often lost in favour of generalising the data points. Therefore, the descion was made to
use a scatter plot with lines intersecting the points to display the change in gradient of the line for different time periods.   

### Display All Data Points
```{r Graph Of Raw Data, echo = FALSE, fig.height = 20, fig.width = 25}
ggplot(data = cancerReg, aes(x = year, y = measure, col = area_name)) +
    geom_smooth(method = "loess", se = FALSE) +
    theme(

        legend.legend.box.spacing = unit(5.0,"cm"),
        legend.title = element_text(size=25),
        legend.text=element_text(size=30),
        legend.key.size = unit(2.0,"cm"),
        axis.text.x = element_text(size = 30),
        axis.text.y = element_text(size = 30), 
        axis.title = element_text(size = 40)
    )
```
The function *geom_smooth()* was used to account for the overplotting that occured for the majority of the healthboards and time periods. This function was paticulary useful as it 
creates smoother lines between the physical data points plotting the measure a specific health board registered for that paticular year. As seen in the plot, most of the time healthboards
will constitently register within ± 20 cases, causing significant overplotting for the multiple healtboards that didnt exhibt drastic changes. However, this function proves very useful


### Find Average of All Measures by Year
```{r Average of All HealthBoards}
avgYearly <- cancerReg %>%
    group_by(year) %>%
    mutate(AvgYear = mean(measure, na.rm = TRUE)) %>%
    select(-area_name, -measure, -area_code)
```

### Calculate Moving Average for Each Health Board
```{r Calculate Moving Average For All Health Boards}
movingAvg <- cancerReg %>%
    group_by(area_name) %>%
    arrange(year) %>%
    mutate(MA = cumsum(measure) / row_number())

```

### Find Last Data Points for Data
```{r Return Last Data Points in Moving Average}
finalValues <- movingAvg %>%
    group_by(area_name) %>%
    summarise(
        lastMA = dplyr::last(MA),
        lastYear=dplyr::last(year)
    )
```




### Display Summary of All Data
```{r Graph of Calculated Averages}
    options(repr.plot.width = 60, repr.plot.height =5) 
    ggplot(data = cancerReg, aes(x = year)) +
        geom_pointline(data = avgYearly, aes(y = AvgYear)) +
        geom_line(data = movingAvg, aes(y = MA, col = area_name)) +
        geom_text_repel(
            data = finalValues, aes(
                x = lastYear,
                y = lastMA,
                label = area_name,
                color = area_name
            ),
            size = 2.5,
            fontface = "bold",
            nudge_x = 5.6,
            direction = "y",
            hjust = 0.7,
            segment.linetype = 2,
            segment.size = 0.5,
            segment.curvature = 0,
            max.overlaps = Inf
        ) +

        theme(legend.position = "none", plot.margin = margin(2,2,2,2))

```

### Calculate Differences Function
```{r Calculate Current Point Against Moving Average}

    sigPercent <- data.frame(
        area_name = character(),
        year = integer(),
        percentNum = numeric(),
        stringsAsFactors = FALSE
    )
    boardAvg <- function(currBoard, currVal, currYear) {
        currAvgYear <- filter(movingAvg, area_name == currBoard & year == currYear) %>% select(MA)
        numCurrAvgYear <- gsub("[^0-9.]", "", currAvgYear$MA)
        numCurrAvgYear <- as.numeric(numCurrAvgYear)
        diffVal <- currVal - numCurrAvgYear
        percentVal <- ((diffVal / numCurrAvgYear) * 100)

        if (percentVal >= 3 || percentVal <= -3) {
            sigPercent <- sigPercent %>% add_row(area_name = currBoard, year = currYear, percentNum = round(percentVal, 2))
        }
        return(sigPercent)

        # WHile the moving average does not provide a direct estimation of the predictied values it still serves as indicator as to what the vale would have looked like in that time frame, and a large deviation from that number means there was a change in data at that time
    }

```

### Calculate Differences

```{r Calculate Differences from Current Point to Moving Average}

    healthBoards <- unique(cancerReg$area_name)
    totalYears <- unique(cancerReg$year)
    for (currBoard in healthBoards) {
        for (currYear in totalYears) {
            currVal <- subset(cancerReg, year == currYear & area_name == currBoard)
            currVal <- select(currVal, -area_code, -area_name, -year)
            currVal <- as.numeric(currVal)
            sigPercent <- boardAvg(currBoard, currVal, currYear)
        }
    }
    sigPercent <- sigPercent %>% arrange(desc(year))
   

``` 

```{r Create Table of Results}

# inputFile <- "reportReg.pdf"
# sigPercentWide_colored <- sigPercentWide %>%
#     mutate(across(everything(), ~ cell_spec(.,
#                                              color = ifelse(. < 0, "red", "black"),
#                                              background = ifelse(. < 0, "lightpink", "white"))))


sigPercentWide <- sigPercent %>% pivot_wider(
    names_from = year,
    values_from = percentNum
)
colnames(sigPercentWide)
sigPercenFinal <- rbind(sigPercentWide, data.frame(area_name = "Total", t(colSums(sigPercentWide[,-1]))))
# sigPercentWide <- sigPercentWide %>% mutate(
#   across(-1,

#       ~ cell_spec(., color = ifelse(is.na(.), "black", ifelse(. < 0, "green", "red")))
#   )
#   )
# sigPercentWide <- sigPercentWide %>% mutate(across(everything(), ~ replace_na(., 0)))
kable(sigPercentWide, format = "latex", booktabs = TRUE, longtable=TRUE) %>%
    kable_styling(latex_options = c("striped", "hold_postion")) %>%
    row_spec(0, bold = TRUE) %>%
     kableExtra::landscape()



# qpdf::pdf_rotate_pages(inputFile, pages = 4, angle = 90)

``` 


#Note for next time: what I want to do at this point is to show the changing colours as a difference change if its only within a small amount of chaning values then ignore the calues and do not 
#colour the cell, otherwise red fir a rise and green for a fall 
#<!-- Since there are a lot of data points, some have more varying level of differnces, a difference is calulated between the current cummulatve avergae and the current year data point
#to hilight the differences in which years contributed to a change in the value of the average by varying degrees, wether that be an increase or a decrease -->