--- 
title: "Title Placeholder"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load in Specific Packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(here)
library(lemon)
library(kableExtra)
library(ggplot2)
library(reshape())
library(hexbin)
library(data.table)
library(GGally)
library(formattable)
library(viridis)
library(TTR)
library(zoo)
library(ggrepel)
library(grid)
```

### Load in the data 
```{r}
cancerReg <- read.csv("C:\\Users\\romin\\ToyRepo\\Models\\cancerReg.csv") 
```


### Remove Uncessary Data for Analysis
```{r}
cancerReg <- cancerReg %>% select(-period,-area_type,-type_definition,-indicator, -upper_confidence_interval, -lower_confidence_interval, -numerator)
```

### Display All Data Points
```{r echo = FALSE, fig.height = 28, fig.width = 20}
ggplot(data=cancerReg, aes(x=year, y=measure, col=area_name)) + geom_smooth(method="loess",se=FALSE)
```


### Find Average of All Measures by Year
```{r}
avgYearly <- cancerReg %>%
    group_by(year) %>%
    mutate(AvgYear = mean(measure, na.rm = TRUE)) %>%
    select(-area_name, -measure, -area_code)
```

### Calculate Moving Average for Each Health Board
```{r}
movingAvg <- cancerReg %>%
    group_by(area_name) %>%
    arrange(year) %>%
    mutate(MA = cumsum(measure) / row_number())

```

### Find Last Data Points for Data
```{r}
finalValues <- movingAvg %>%
    group_by(area_name) %>%
    summarise(
        lastMA = dplyr::last(MA),
        lastYear=dplyr::last(year)
    )
```




### Display Summary of All Data
```{r}

ggplot(data=cancerReg, aes(x = year)) +
    geom_pointline(data= avgYearly, aes(y = AvgYear)) +
    geom_pointline(data=movingAvg, aes(y=MA, col=area_name))+
    geom_text_repel(data = finalValues, aes(
        x = lastYear,
        y = lastMA,
        label = area_name,
        color=area_name),
        size = 2.5, 
        fontface = "bold",
        nudge_y = 20.6,
        direction = "y",
        hjust= -0.7,
        segment.linetype=2,
        segment.size = 0.5,
        segment.curvature=0
        ) +
    theme(legend.position = "none") 

```

### Calculate Differences Function
```{r}
boardAvg <- function(currBoard, currVal,currYear){
 
    currAvgYear <- filter(movingAvg, area_name == currBoard & year == currYear) %>% select(MA)
    numCurrAvgYear <- gsub("[^0-9.]", "", currAvgYear$MA)
    numCurrAvgYear <- as.numeric(numCurrAvgYear)
    diffVal <- currVal - numCurrAvgYear
    percentVal <- ((diffVal / numCurrAvgYear) * 100)
    
    if (percentVal >=3 || percentVal <=-3){
       print(currBoard)
       print(currYear)
       print(percentVal)
       print("---------------------")
    } 


## Turn these differences into a percentage of how different it is, and if its greatly different return back the percentage value

}
```

### Calculate Differences
<!-- Since there are a lot of data points, some have more varying level of differnces, a difference is calulated between the current cummulatve avergae and the current year data point
to hilight the differences in which years contributed to a change in the value of the average by varying degrees, wether that be an increase or a decrease -->
```{r}
healthBoards <- unique(cancerReg$area_name)
totalYears <- unique(cancerReg$year)
for (currBoard in healthBoards){
    for (currYear in totalYears){
        currVal <- subset(cancerReg, year == currYear & area_name == currBoard)
        currVal <- select(currVal, -area_code, -area_name, -year)
        currVal <- as.numeric(currVal)
        boardAvg(currBoard, currVal, currYear)
       

    }
       
}

``` 

```{r}
colourCells <- function(values, average){
    diffVal <- values - average
    if (diffVal>=0 & diffVal<=20){
        return(paste0("\\cellcolor{green!," ,round(diffVal/100), "}"))
    }
}
```



### Summary Table of Data Within Graph 
```{r}
inputFile <-"reportReg.pdf"
healthBoardData <- cancerReg %>%
    group_by(year) %>%
    mutate(Percent = measure / sum(measure) * 100)
healthBoardData <- cancerReg %>% pivot_wider(names_from = year, values_from = measure)
healthBoardData <- healthBoardData %>%
    mutate(
        Average = rowMeans(select(., `2003`:`2020`), na.rm = TRUE)
    )

kable(healthBoardData, format = "latex", booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "hold_poistion")) %>%
    row_spec(0, bold = TRUE) %>%
    kableExtra::landscape()
qpdf::pdf_rotate_pages(inputFile, pages=4 , angle=90)
``` 


#Note for next time: what I want to do at this point is to show the changing colours as a difference change if its only within a small amount of chaning values then ignore the calues and do not 
#colour the cell, otherwise red fir a rise and green for a fall 