--- 
title: "Title Placeholder"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load in Specific Packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(here)
library(lemon)
library(kableExtra)
library(ggplot2)
library(reshape())
library(hexbin)
library(data.table)
library(GGally)
library(formattable)
library(viridis)
library(TTR)
library(zoo)
library(ggrepel)
```

### Load in the data 
```{r}
cancerReg <- read.csv("C:\\Users\\romin\\ToyRepo\\Models\\cancerReg.csv") 
```


### Remove Uncessary Data for Analysis
```{r}
cancerReg <- cancerReg %>% select(-period,-area_type,-type_definition,-indicator, -upper_confidence_interval, -lower_confidence_interval, -numerator)
```


### Display Summary of All Data
```{r}

# ggplot(cancerReg,aes(x=year, y= measure, col=area_name)) + geom_line() + geom_point()
# ggplot(cancerReg, aes(x=area_name, y=measure, fill=area_name)) + geom_violin()
# ggparcoord(data= cancerReg, columns = c(), groupColumn = "area_name")
# ggplot(cancerReg, aes(fill=year, y=measure, x=area_name)) + geom_bar(position = "stack", stat = "identity")
# ggplot(cancerReg, aes(x = year, y = measure, fill = area_name)) +  geom_area(alpha=0.6 , size=.5, colour="white") +
#     scale_fill_viridis(discrete = T)
# ggplot(cancerReg, aes(x = year, y = measure, col = area_name)) +
#     geom_point(size = 3, position = position_jitter(h = 0.50, w = 0.50), alpha = 0.5) +
#     geom_smooth(se=FALSE, size = 0.7 )


cancerReg <- cancerReg %>%
    arrange(year) %>%
    mutate(MA = rollmean(cancerReg$measure, k = 14, fill = NA, align = "right"))

subData <- cancerReg[seq(1, nrow(cancerReg), 14), ]
ggplot(cancerReg, aes(x = year)) +
    geom_smooth(aes(y = measure, col = area_name), se = FALSE, size = 0.5, alpha = 0.5) +
    # geom_text_repel(data = cancerReg, aes(y = measure, label = area_name), nudge_x = 1, na.rm = TRUE)+
    geom_text_repel(aes(colour=area_name))
    geom_point(data = subData, aes(y = MA), color = "black", size = 2) +
    geom_line(data = subData, aes(y = MA), colour = "black", linetype = "dashed", size = 0.7) +
    theme(legend.position = "none")
    


``` 

```{r}
colourCells <- function(values, average){
    diffVal <- values - average
    if (diffVal>=0 & diffVal<=20){
        return(paste0("\\cellcolor{green!," ,round(diffVal/100), "}"))
    }
}
```



### Summary Table of Data Within Graph 
```{r}
healthBoardData <- cancerReg %>%
    group_by(year) %>%
    mutate(Percent = measure / sum(measure) * 100)
healthBoardData <- cancerReg %>% pivot_wider(names_from = year, values_from = measure)
healthBoardData <- healthBoardData %>%
    mutate(
        Average = rowMeans(select(., `2003`:`2020`), na.rm = TRUE)
    )
kable(healthBoardData, format = "latex", booktabs = TRUE)
healthBoardData
``` 


#Note for next time: what I want to do at this point is to show the changing colours as a difference change if its only within a small amount of chaning values then ignore the calues and do not 
#colour the cell, otherwise red fir a rise and green for a fall 